### 2008-03-18 trunk/Papers/dl-gre-08/conclusion.tex
added structure and some references
wrote first versions of intro, related work, efficiency
iterated 4.3 and 5; updated runtimes picture

\section{Computing bisimulation classes} \label{sec:bisim}

>>>>

\section{Conclusion} \label{sec:conclusion}

referring to sets (or is this trivial?)

where do we talk about mutually constraining descriptions like ``the
rabbit in the hat''?

will make implementation available

first time to actually show efficient runtimes for generation of
relational REs

Because they compute referring expressions for all individuals in the
domain together, our algorithms will be strongest in static settings,
such as the generation of descriptions for museum exhibits, in which
the individuals and their properties don't change much.  Nevertheless,
they remain fast enough for real-time use in dynamic settings.
Another advantage that recent algorithms for relational REs
\cite{Krahmer2003,kelleher06:_increm_gener_of_spatial_refer} have over
ours is that they can take weights and salience into account.  How
this could be incorporated into our algorithms is an interesting
question for future research.

relational descriptions in data are comparatively rare and
inconsistent; not sufficiently annotated in TUNA
\cite{deemter06:_build_seman_trans_corpus_for}; need more.

Things that we care about but that are never said explicitly so far:

\begin{itemize}
\item in a static scenario, we only have to compute REs once -- and in
  any application that uses many REs before the context changes (e.g.\
  museums), this is exactly what we want to do
\item it is difficult to control the search for concepts by linguistic
  constrtaints
\item our GRE algorithms can take orderings over the propositional and
  relational symbols into account
\end{itemize}

### 2008-03-18 trunk/Papers/dl-gre-08/discussion.tex
iterated 4.3 and 5; updated runtimes picture

%evaluate on TUNA \cite{deemter06:_build_seman_trans_corpus_for} or a
%richer version of it

%or on the cabinets corpus
%\cite{viethen06:_algor_for_gener_refer_expres}

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{runtimes}
  \caption{Average runtimes (in ms) of the two algorithms on random graphs of
    different sizes.} 
  \label{fig:runtimes}
\end{figure}

In order to get a more comprehensive picture of our system's
efficiency, we ran the two algorithms on random graphs with increasing
numbers of nodes.  \todo{harmonize this with what Carlos calls them in
  Section~\ref{sec:bisim}} Each graph had ten different atoms and four
different roles, a 10\% chance for a node to have each atom, and a
10\% chance for each pair of nodes to be connected by each role.  The
results (averaged over 10 runs for each graph size) are shown in
Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\ algorithm
takes about 350 ms on average to generate relational REs for all
individuals in the graph with 100 nodes, i.e.\ 4 ms on average for
each individual.  The \alc\ algorithm is even faster, at about 200 ms
for the graph with 100 nodes.  In both cases, we conclude that the
systems are fast enough for practical use.

@@

Like other algorithms that do content determination without taking
into account linguistic contraints, we have no guarantee that the
descriptions generated by our algorithm can actually be realized. For
example, none of the descriptions out algorithms generate in the
drawer domain contain a property which would commonly be realized as a
noun (the property \textsf{drawer} is not included because it applies
to all entities in this domain). If we fed this output directly into a
realizer, the realizer would thus be faced with the task of having to
produce a noun phrase without being given anything that could serve as
the head noun. Of course, this particular problem could easily be
dealt with through a post-processing step that adds a type property to
the description. But at least for the \alc\ algorithm it is not clear
that all non-realizable descriptions could be fixed through such
simple post-processing solutions.

Partly in response to this problem, Horacek \shortcite{} proposes to
interleave content determination with realization and build the
linguistic representation of a natural language description in
parallel with building the representation of its semantic content.

SPUD \cite{} is another system that avoids this problem by
interleaving content determination with realization. SPUD goes one
step further and instead of generating referring expressions in
isolation, deals with them in the context of the sentence they appear
in taking into account semantic information contributed by other parts
of that sentence. This allows them to generate utterances like
\textit{``remove the rabbit from the hat''} which can be used in contexts
containing several rabbits and several hats as long as there is only
one rabbit-hat pair where the rabbit is in the hat. While our
algorithm does generate referring expressions for all entities in the
context at once (and thereby avoids the problem of infinite regress
\todo{discussed earlier}), it is currently not designed to take into
account additional information provided by the context. When used to
generate the expressions referring to the relevant rabbit and hat in
the situation described above, our algorithms would produce
\textit{``the rabbit in the hat''} and \textit{``the hat with the
rabbit''}, respectively, yielding \textit{``remove the rabbit in the
hat from the hat with the rabbit''}.

To improve the interface between our algorithm and a realization
component, we therefore need to investigate ways of how linguistic
contraints on realization can be used to guide our content
determination algorithms and how it can be integrated into the process
of generating a whole utterance in such a way that relevant
information contributed by other parts of the utterance can be taken
into account.

>>>>

@@

\begin{figure}
  \centering
  \includegraphics[width=\columnwidth]{runtimes}
  \caption{Average runtimes (in ms) of the two algorithms on random
    models of     different sizes.} 
  \label{fig:runtimes}
\end{figure}

In order to get a more comprehensive picture of the algorithms'
efficiency, we ran them on random models with increasing numbers of
individuals.  Each model had random interpretations for ten different
propositional and four relational symbols; each individual had a 10\%
chance to be in the extension of each propositional symbol, and each
pair of individuals had a 10\% chance to be related by a relational
symbol.  The results (averaged over 10 runs for each model size) are
shown in Fig.~\ref{fig:runtimes}.  As the figure shows, the \el\
algorithm takes about 350 ms on average to generate relational REs for
all individuals in the model of size 100, i.e.\ less than 4 ms on
average for each individual.  The \alc\ algorithm is even faster, at
about 140 ms for the model of size 100.  As far as we know, these are
by far the fastest published runtimes for any relational GRE algorithm
in the literature.

Another encouraging aspect of both runtime graphs is that they seem to
be approximated well by quadratic functions.  So although the
worst-case runtime of the \el\ algorithm might be exponential, both
algorithms empirically exhibit (low) polynomial runtime on the inputs
we considered.

@@

Our GRE algorithms do not guarantee that the concepts they compute can
actually be realized in language.  For example, none of the concepts
our algorithms computed in the Viethen \& Dale domain contained an
atom that would commonly be realized as a noun; the property
$\mathsf{drawer}$ is never used because it applies to all individuals
in the domain.  This particular problem could easily be worked around
in a post-processing step.  However, another problem arises from the
massive use of negation in the \alc\ algorithm; it will be hard for
any realizer to find a reasonable way of expressing the concept $\neg
\exists R.(\neg P \sqcap \neg Q)$ from the example in
Fig.~\ref{fig:el-vs-alc} as a smooth noun phrase.  Although we agree
with \newcite{deemter01:_gener_refer_expres} and others that the
careful use of negation and disjunction can improve REs, these
connectives must not be overused.  Thus we consider the concepts
computed by the \el\ algorithm ``safer'' with respect to realization.

Of course, we share the problem of interfacing GRE and realization
with every other approach that separates these two modules, i.e.\
almost the entire GRE literature (notable exceptions are e.g.\ SPUD
\cite{Stone2003a} and \newcite{Horacek1997}).  In principle, we
believe that it is a good idea to handle sentence planning and
realization in a single module; for instance, SPUD can use its
awareness of the syntactic context to generate succinct REs as in
``take the rabbit from the hat''.  We hope that the ideas we have
explored here for efficient and expressive RE generation can
eventually be combined with recent efficient algorithms for integrated
sentence planning and realization, such as in \newcite{KolSto07}.

One problem that arises in our approach specifically is that the
bisimulation classes algorithms derive some measure of efficiency from
their freedom to build concepts without having to respect any
linguistic constraints.  It seems straightforward to extend Krahmer et
al.'s \shortcite{Krahmer2003} approach such that it only considers
subgraphs that can actually be realized, because their algorithm
proceeds by a genuine search for uniquely identifying subgraphs, and
will simply take a different branch of the search if some subgraph is
useless.  Our algorithms don't search in the same way; any constraints
that disallow certain refinements of a partition jeopardize the
algorithms' ability to find a partition into singletons in polynomial
time.  A particularly vivid example is the fact that both algorithms
crucially split over all propositional symbols before they start
iterating over relational symbols; changing this may make them
incomplete.  Investigating this interplay between efficiency and
linguistic constraints is an interesting avenue for future research.

### 2008-03-18 trunk/Papers/dl-gre-08/related.tex
explained my algorithm
iterated 4.3 and 5; updated runtimes picture

\section{Introduction} \label{sec:introduction}

\cite{Dale1995}
\cite{dale91:_gener_refer_expres_invol_relat}

>>>>

\section{A unified perspective on GRE} \label{sec:related}

\begin{figure}
  \centering
  \begin{tabular}{l|l}
    GRE algorithm & DL variant \\ \hline
    \newcite{Dale1995} & only conjunction \\
    \newcite{deemter01:_gener_refer_expres} & propositional connectives \\
    \newcite{dale91:_gener_refer_expres_invol_relat} & \el \\
    \newcite{Krahmer2003} & \el \\
    \newcite{kelleher06:_increm_gener_of_spatial_refer} & \el \\
    \newcite{gardent02:_gener_minim_defin_descr} & \alc
  \end{tabular}
  \caption{DL variants used by different GRE algorithms.}
  \label{fig:related}
\end{figure}

The view of GRE as a problem of computing DL concepts allows us to
organize existing approaches to GRE with respect to the logical
connectives of DL which they can use in the referring expressions they
generate.  We have summarized this for some approaches in
Fig.~\ref{fig:related}.  We already mentioned some of these in
Section~\ref{sec:gre} already.  In addition, the graphs
\newcite{Krahmer2003} use to represent referring expressions can be
seen as concepts of \el\ that are only satisfied at those points at
which they can be embedded, and the descriptions generated by
\newcite{gardent02:_gener_minim_defin_descr} are all concepts of \alc,
without requiring that the bisimulation class must be singleton (to be
able to refer to sets).

Next to the inventory of DL connectives, we can also compare the
algorithms that different approaches use to compute the concepts.  In
a purely propositional setting, where there are no relations, our \el\
algorithm performs essentially the same computation as the Incremental
Algorithm \cite{Dale1995}, i.e.\ split the domain according to the
positive atoms.  However, in the relational case, our algorithms
operate very differently than any of the ones listed in
Fig.~\ref{fig:related}.  Rather than attempting to generate a RE for a
single individual and then doing the same for their neighbors if
necessary, with all the associated risks of infinite regress that this
entails (and that then have to be avoided by additional heuristics),
we successively refine a decomposition of the domain into smaller
classes, effectively computing REs for all individuals at the same
time.  That is, our algorithm is not a search algorithm in the sense
of \newcite{DBLP:conf/ijcai/BohnetD05}, and at least the \alc\
algorithm can therefore compute concepts of minimal relational depth
in worst-case polynomial time.  Perhaps closest in spirit is the
Krahmer et al.\ graph algorithm, which also computes \el\ concepts by
extending them successively; but this algorithm, too, is focused on
reference to a single individual, and it has NP-complete worst-case
complexity.
